---
title: "Quantium Virtual Internship - Retail Strategy and Analytics - Task 1"
output:
  html_document: default
  pdf_document: default
  word_document: default
---


```{r setup, include=FALSE}
# set options for R markdown knitting
knitr::opts_chunk$set(echo = TRUE)
```

## Solution template for Task 1

This file is a solution template for the Task 1 of the Quantium Virtual Internship. It will walk you through the analysis, providing the scaffolding for your solution with gaps left for you to fill in yourself.
Look for comments that say "over to you" for places where you need to add your own
code!

Often, there will be hints about what to do or what function to use in the text leading up to a code block - if you need a bit of extra help on how to use a function, the internet has many excellent resources on R coding, which you can find using your favourite search engine.

## Load required libraries and datasets
Note that you will need to install these libraries if you have never used these before.

```{r cars}
setwd("~/WORK/Forage/quantium/Task_1_Data preparation and customer analytics")
```

```{r}
#### Example code to install packages
#install.packages("data.table")
```

```{r}
#### Load required libraries
library(data.table)
library(ggplot2)
library(ggmosaic)
library(readr)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}

```

